Constraints on quasi--normal-mode frequencies with LIGO-Virgo binary--black-hole observations

### 2021-03-25 papers/intro_paper/acknowledgements.tex
minor changes through the text and added some comments

The authors would like to thank everyone at the frontline of the Covid-19 pandemic. We would like to acknowledge the computing resources at the AEI and the LIGO Scientific Collaboration, specifically the Hypatia and CIT clusters. R.B. acknowledges financial support from the European Union's Horizon 2020 research and innovation programme under the Marie Sk\l odowska-Curie grant agreement No. 792862, from the European Union's H2020 ERC, Starting Grant agreement no.~DarkGRA--757480, and from the MIUR PRIN and FARE programmes (GW-NEXT, CUP:~B84I20000100001). We also acknowledge support from the Amaldi Research Center funded by the MIUR program ``Dipartimento di Eccellenza'' (CUP:~B81I18001170001).

>>>>

\rb{The authors thank Max Isi for generously allowing us to use the code developed in Ref.~\cite{Isi:2019asy} for the hierarchical inference analysis.} The authors would like to thank everyone at the frontline of the Covid-19 pandemic. We would like to acknowledge the computing resources at the AEI and the LIGO Scientific Collaboration, specifically the Hypatia and CIT clusters. R.B. acknowledges financial support from the European Union's Horizon 2020 research and innovation programme under the Marie Sk\l odowska-Curie grant agreement No. 792862, from the European Union's H2020 ERC, Starting Grant agreement no.~DarkGRA--757480, and from the MIUR PRIN and FARE programmes (GW-NEXT, CUP:~B84I20000100001). We also acknowledge support from the Amaldi Research Center funded by the MIUR program ``Dipartimento di Eccellenza'' (CUP:~B81I18001170001).

### 2021-03-25 papers/intro_paper/appendix.tex
minor changes through the text and added some comments

Inferences of all parameters in this paper have been done under the
assumption that the noise in the detectors is stationary and
Gaussian. In other words, detector noise follows a normal distrubution
with zero mean and a PSD, $S_n(f)$, that is not a function of time, at
least during the internal of the GW signal. This
allows us to write the Bayesian likelihood function in the form given
in Eqs.~(\ref{eq:likelihood}) and ~(\ref{eq:nwip}), and perform all the
parameter estimation that follows in the results sections. However,,
LIGO-Virgo noise can often have features that deviate from
stationarity and Gaussianity. If such features are not taken into
account appropriately, final estimates of parameters can get
biased. Here we demonstrate one such case by injecting in real noise a
GW190521-like signal and showing how parameter estimates can be biased
when our description of detector noise is not complete.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\begin{center}
        \includegraphics[width=0.4\textwidth]{figures/S190521g_swinjs.png}
        \caption{\textcolor{red}{FINAL RESULT} 90 \% credible level on the posterior probability distribution of the frequency and damping time of $(2,\pm 2)$ mode, $(\fgr{220}, \taugr{220})$ using synthetic \texttt{NRSur} signals with parameters similar to the GW event, GW190521, in Gaussian noise (grey dot-dashed lines) and real interferometric noise (green dot dashed lines). The GR prediction for the frequency and damping time is indicated by the black cross. While the Gaussian noise simulations are consistent with the prediction, at least 3 or the 5 real noise simulation are not. The black curve corresponds to the measurements of the real event GW190521 reported in Ref.~\cite{Abbott:2020jks}. All signals are recovered using the $\pSEOB$ model.}
        \label{fig:21g_systematics}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

To justify this statement, in Fig.~\ref{fig:correlations} we show corner plots that illustrate the correlations between the non-GR parameters $(\df{220}, \dtau{220})$, the detector-frame total mass $M(1+z)$ and the reconstructed frequency and damping time $(f_{220},\tau _{220})$ for GW150914 (left panel), corresponding to an event for which both the pre- and post-merger phase are measurable, and for GW190521 (right panel), an event in which the post-merger has ${\rm SNR}> 8$ but the pre-merger has an SNR below 8. For GW190521, due the strong degeneracy between $\df{220}$ and the $M(1+z)$, the 1D posterior for $M(1+z)$ is pushed towards the upper boundary of its prior, despite the very wide prior employed in the analysis. This in turn renders the measurement of $\df{220}$ (and to a smaller degree of $\dtau{220}$) highly dependent on the upper boundary of the total mass prior. On the other hand, this issue does not significantly affect the posteriors for the reconstructed quantities $(f_{220},\tau _{220})$ which are well measured and nearly independent on the upper prior boundary for $M(1+z)$. 
%
This is to be contrasted with the results for GW150914. In this case, the extra information coming from the pre-merger phase allows to break the degeneracy between the non-GR parameters and the total mass, and therefore both $(\df{220}, \dtau{220})$ and $M(1+z)$ can be measured at the same time.

>>>>

Inferences of all parameters in this paper have been done under the
assumption that the noise in the detectors is stationary and
Gaussian. In other words, detector noise follows a normal distribution
with zero mean and a PSD, $S_n(f)$, that is not a function of time, at
least during the duration of the GW signal. This
allows us to write the Bayesian likelihood function in the form given
in Eqs.~(\ref{eq:likelihood}) and ~(\ref{eq:nwip}), and perform all the
parameter estimation that follows in the results sections. However,
LIGO-Virgo noise can often have features that deviate from
stationarity and Gaussianity. If such features are not taken into
account appropriately, final estimates of parameters can get
biased. Here we demonstrate one such case by injecting in real noise a
GW190521-like signal and showing how parameter estimates can be biased
when our description of detector noise is not complete.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\begin{center}
        \includegraphics[width=0.4\textwidth]{figures/S190521g_swinjs.png}
        \caption{\textcolor{red}{FINAL RESULT} 90 \% credible level on the posterior probability distribution of the frequency and damping time of $(2,\pm 2)$ mode, $(\fgr{220}, \taugr{220})$ using synthetic \texttt{NRSur} signals with parameters similar to the GW event, GW190521, in Gaussian noise (grey dot-dashed lines) and real interferometric noise (green dot dashed lines). The GR prediction for the frequency and damping time is indicated by the black cross. While the Gaussian noise simulations are consistent with the prediction, at least 3 of the 5 real noise simulation are not. The black curve corresponds to the measurements of the real event GW190521 reported in Ref.~\cite{Abbott:2020jks}. All signals are recovered using the $\pSEOB$ model.}
        \label{fig:21g_systematics}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

To justify this statement, in Fig.~\ref{fig:correlations} we show corner plots that illustrate the correlations between the non-GR parameters $(\df{220}, \dtau{220})$, the detector-frame total mass $M(1+z)$ and the reconstructed frequency and damping time $(f_{220},\tau _{220})$ for GW150914 (left panel), corresponding to an event for which both the pre- and post-merger phase are measurable, and for GW190521 (right panel), an event in which the post-merger has ${\rm SNR}> 8$ but the pre-merger has an SNR below 8. For GW190521, due the strong degeneracy between $\df{220}$ and $M(1+z)$, the 1D posterior for $M(1+z)$ is pushed towards the upper boundary of its prior, despite the very wide prior employed in the analysis. This in turn renders the measurement of $\df{220}$ (and to a lesser degree of $\dtau{220}$) highly dependent on the upper boundary of the total mass prior. On the other hand, this issue does not significantly affect the posteriors for the reconstructed quantities $(f_{220},\tau _{220})$ which are well measured and nearly independent on the upper prior boundary for $M(1+z)$. 
%
This is to be contrasted with the results for GW150914. In this case, the extra information coming from the pre-merger phase allows to break the degeneracy between the non-GR parameters and the total mass, and therefore both $(\df{220}, \dtau{220})$ and $M(1+z)$ can be measured at the same time.

### 2021-03-25 papers/intro_paper/discussion.tex
minor changes through the text and added some comments

\ab{We have built a parameterized IMR waveform model, $\pSEOB$, that can 
measure the QNM complex frequencies of the remnant object formed \ab{through 
the merger of BHs with aligned or antialigned spins, thus extending  
previous work, which was limited} to non-spinning BHs~\cite{Brito:2018rfr}. 
The $\pSEOB$ model was recently used to infer the QNMs of some of the 
BBH's remnants detected by LIGO and Virgo during O3a~\cite{Abbott:2020jks}.}

\ab{After testing our method to infer the QNM frequencies with GR and non-GR synthetic-signal 
injections in Gaussian noise, we have applied it to LIGO and Virgo real data. 
We have analyzed GW events in O1, O2 and O3a (a total of 4 new events) that were not examined in Ref.~\cite{Abbott:2020jks} 
with the $\pSEOB$ model (see Table~\ref{tab:qnm_o1o2_results}). After combining our new 
results with the other GW events in O3a investigated with our method~\cite{Abbott:2020jks}, 
we have obtained more stringent bounds on the dominant (or least-damped) QNM $(\ell=2,m=2)$.}

\ab{More specifically, as expected, the single GW event providing the best constraint
on the least-damped QNM to date continues to be GW150914 $(\df{220}=0.04^{+0.06}_{-0.04}, 
\dtau{220}=0.09^{+0.18}_{-0.18})$. In addition, in the most agnostic and
conservative scenario where we combine the information from different
events using a hierarchical approach~\cite{Zimmerman:2019wzo,Isi:2019asy}, we obtain 
at $90\%$ credibility $(\df{220}=0.02^{+0.09}_{-0.09}, \dtau{220}=0.13^{+0.42}_{-0.40})$. Thus, 
our results constrain the frequency (decay time) of the least damped QNM to be within $\sim
9\%$ ($\sim 40\%$) of the GR prediction --- an improvement of a factor of $\sim 4$ ($\sim 2$) 
over the results obtained with the $\pSEOB$ model in Ref.~\cite{Abbott:2020jks}.}

@@

\ab{These constraints could in principle be used to constrain specific
non-GR theories and exotic compact objects~\cite{Glampedakis:2017cgd,Cardoso:2019rvt,Maggio:2020jml}. 
However, QNM computations in non-GR theories have only been done in an
handful of cases, mostly focusing on non-rotating or slowly-rotating
BH solutions~\cite{Ferrari:2000ep,Molina:2010fb,Pani:2009wy,Blazquez-Salcedo:2016enn,Blazquez-Salcedo:2017txk,Brito:2018hjh,Franciolini:2018uyq,Cardoso:2018ptl,Tattersall:2018nve,Tattersall:2019nmh,Blazquez-Salcedo:2019nwd,Silva:2019scu,Glampedakis:2019dqh,Blazquez-Salcedo:2020jee,Blazquez-Salcedo:2020caw,Cano:2020cao}, 
or relying on the eikonal/geometric optics approximation to obtain
estimates of the QNMs for spinning BHs~\cite{Blazquez-Salcedo:2016enn,Glampedakis:2017dvb,Jai-akson:2017ldo}. The
only exceptions to this rule, that we are aware of, are computations
of the QNM of Kerr-Newman BHs in Einstein-Maxwell
theory~\cite{Pani:2013ija,Pani:2013wsa,Mark:2014aja,Dias:2015wqa} or 
estimates of the BBH in non-GR theories obtained through a limited number of 
NR simulations~\cite{Okounkova:2019dfo,Okounkova:2019zjf}. Given these
limitations, our ability of going beyond a null test of GR and use our
results to impose precise constraints on non-GR theories with QNM
measurements is currently quite limited.}

Despite these \ab{theoretical} limitations there has been some recent effort to develop
parametrizations that could help mapping measurements of the
parameters $(\delta f_{\ell m 0}, \delta \tau_{\ell m0})$ onto
constraints to specific non-GR theories. This includes for example
the parametrization proposed in Ref.~\cite{Maselli:2019mjd}, recently
applied to LIGO-Virgo GW events in Ref.~\cite{Carullo:2021dui}, where
deviations from the GR QNMs explicitly depend on a perturbative
expansion in the BH spin and possible extra non-GR parameters, or the
proposal of Refs.~\cite{Cardoso:2019mqo,McManus:2019ulj} where
deviations from the GR QNMs are mapped onto generic small
modifications of the perturbation equations describing the QNMs. Other
examples also include proposals to map deviations from the GR QNMs to
coefficients in generic effective--field-theory
actions~\cite{Cardoso:2018ptl,Franciolini:2018uyq,Cano:2020cao} or to
directly relate measurements of the QNM complex frequencies to a
parametrized non-GR BH metric~\cite{Glampedakis:2017dvb,Suvorov:2021amy,Volkel:2020daa}, 
which could be then used jointly with measurements from the Event
Horizon Telescope to obtain stronger constraints on deviations from
GR~\cite{Volkel:2020daa,Volkel:2020xlc,Psaltis:2020lvx,Yang:2021zqy}.
We should note, however, that all these parametrizations are either
limited to non-spinning BHs or make use of a series expansion in the
BH spin which might limit their accuracy for highly-spinning BHs\ab{, unless 
the sensitivity of GW detectors will not allow us to access the higher 
coefficients in the spin series.}

@@

\ab{Finally, an obvious generalization of this work would be to extend the
parameterized $\pSEOB$ model to generic precessing BBHs, which can in
principle be easily done using the recently developed multipolar EOB
model reported in Ref.~\cite{Ossokine:2020kjp}. Lastly, it will be relevant 
to include GR deviations in the $\pSEOB$ model also during the late inspiral 
and plunge stages of the BBH coalescence. On the other hand, GR deviations (notably 
deviations from PN theory) for the long inspiral stage are currently 
available in the $\pSEOB$ model and have been used to set bounds on the 
PN parameters in the GW phasing using LIGO and Virgo observations~\cite{Abbott:2018lct,
LIGOScientific:2019fpa,Abbott:2020jks}.}

>>>>

We have built a parameterized IMR waveform model, $\pSEOB$, that can 
measure the QNM complex frequencies of the remnant object formed through 
the merger of BHs with aligned or antialigned spins, thus extending  
previous work, which was limited to non-spinning BHs~\cite{Brito:2018rfr}. 
The $\pSEOB$ model was recently used to infer the QNMs of some of the 
BBH's remnants detected by LIGO and Virgo during O3a~\cite{Abbott:2020jks}.

After testing our method to infer the QNM frequencies with GR and non-GR synthetic-signal 
injections in Gaussian noise, we have applied it to LIGO and Virgo real data. 
We have analyzed GW events in O1, O2 and O3a (a total of 4 new events) that were not examined in Ref.~\cite{Abbott:2020jks} 
with the $\pSEOB$ model (see Table~\ref{tab:qnm_o1o2_results}). After combining our new 
results with the other GW events in O3a investigated with our method~\cite{Abbott:2020jks}, 
we have obtained more stringent bounds on the dominant (or least-damped) QNM $(\ell=2,m=2)$.

More specifically, as expected, the single GW event providing the best constraint
on the least-damped QNM to date continues to be GW150914 $(\df{220}=0.04^{+0.06}_{-0.04}, 
\dtau{220}=0.09^{+0.18}_{-0.18})$. In addition, in the most agnostic and
conservative scenario where we combine the information from different
events using a hierarchical approach~\cite{Zimmerman:2019wzo,Isi:2019asy}, we obtain 
at $90\%$ credibility $(\df{220}=0.02^{+0.09}_{-0.09}, \dtau{220}=0.13^{+0.42}_{-0.40})$. Thus, 
our results constrain the frequency (decay time) of the least damped QNM to be within $\sim
9\%$ ($\sim 40\%$) of the GR prediction --- an improvement of a factor of $\sim 4$ ($\sim 2$) 
over the results obtained with the $\pSEOB$ model in Ref.~\cite{Abbott:2020jks}.

@@

These constraints could in principle be used to constrain specific
non-GR theories and exotic compact objects~\cite{Glampedakis:2017cgd,Cardoso:2019rvt,Maggio:2020jml}. 
However, QNM computations in non-GR theories have only been done in an
handful of cases, mostly focusing on non-rotating or slowly-rotating
BH solutions~\cite{Ferrari:2000ep,Molina:2010fb,Pani:2009wy,Blazquez-Salcedo:2016enn,Blazquez-Salcedo:2017txk,Brito:2018hjh,Franciolini:2018uyq,Cardoso:2018ptl,Tattersall:2018nve,Tattersall:2019nmh,Blazquez-Salcedo:2019nwd,Silva:2019scu,Glampedakis:2019dqh,Blazquez-Salcedo:2020jee,Blazquez-Salcedo:2020caw,Cano:2020cao,Wagle:2021tam,Pierini:2021jxd}, 
or relying on the eikonal/geometric optics approximation to obtain
estimates of the QNMs for spinning BHs~\cite{Blazquez-Salcedo:2016enn,Glampedakis:2017dvb,Jai-akson:2017ldo}. The
only exceptions to this rule, that we are aware of, are computations
of the QNM of Kerr-Newman BHs in Einstein-Maxwell
theory~\cite{Pani:2013ija,Pani:2013wsa,Mark:2014aja,Dias:2015wqa} or estimates of the BBH in non-GR theories obtained
through a limited number of NR simulations~\cite{Okounkova:2019dfo,Okounkova:2019zjf}. Given these
limitations, our ability of going beyond a null test of GR and use our
results to impose precise constraints on non-GR theories with QNM
measurements is currently quite limited.

Despite these theoretical limitations there has been some recent effort to develop
parametrizations that could help mapping measurements of the
parameters $(\delta f_{\ell m 0}, \delta \tau_{\ell m0})$ onto
constraints to specific non-GR theories. This includes for example
the parametrization proposed in Ref.~\cite{Maselli:2019mjd}, recently
applied to LIGO-Virgo GW events in Ref.~\cite{Carullo:2021dui}, where
deviations from the GR QNMs explicitly depend on a perturbative
expansion in the BH spin and possible extra non-GR parameters, or the
proposal of Refs.~\cite{Cardoso:2019mqo,McManus:2019ulj} where
deviations from the GR QNMs are mapped onto generic small
modifications of the perturbation equations describing the QNMs. Other
examples also include proposals to map deviations from the GR QNMs to
coefficients in generic effective--field-theory
actions~\cite{Cardoso:2018ptl,Franciolini:2018uyq,Cano:2020cao} or to
directly relate measurements of the QNM complex frequencies to a
parametrized non-GR BH metric~\cite{Glampedakis:2017dvb,Suvorov:2021amy,Volkel:2020daa}, 
which could be then used jointly with measurements from the Event
Horizon Telescope to obtain stronger constraints on deviations from
GR~\cite{Volkel:2020daa,Volkel:2020xlc,Psaltis:2020lvx,Yang:2021zqy}.
We should note, however, that all these parametrizations are either
limited to non-spinning BHs or make use of a series expansion in the
BH spin which might limit their accuracy for highly-spinning BHs\ab{, unless 
the sensitivity of GW detectors will not allow us to access the higher 
coefficients in the spin series.}

@@

Finally, an obvious generalization of this work would be to extend the
parameterized $\pSEOB$ model to generic precessing BBHs, which can in
principle be easily done using the recently developed multipolar EOB
model reported in Ref.~\cite{Ossokine:2020kjp}. Lastly, it will be relevant 
to include GR deviations in the $\pSEOB$ model also during the late inspiral 
and plunge stages of the BBH coalescence. On the other hand, GR deviations (notably 
deviations from PN theory) for the long inspiral stage are currently 
available in the $\pSEOB$ model and have been used to set bounds on the 
PN parameters in the GW phasing using LIGO and Virgo observations~\cite{Abbott:2018lct,
LIGOScientific:2019fpa,Abbott:2020jks}.
